---
title: "An Interactive Tool for Understanding/Quantifying the Variability in Classification Performance Metrics with Class Imbalance"
output: 
  flexdashboard::flex_dashboard:
    theme: 
      version: 4
      bootswatch: simplex
    includes: 
      in_header: header.html
      after_body: busy.html
runtime: shiny
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

pacman::p_load(tidyverse, yardstick, # used for calculations
               plotly, # for interactive plots
               shiny, flexdashboard, DT, # for the shiny/flexdashboard functions
               fontawesome) # for neat looking icons

source('https://raw.githubusercontent.com/fmegahed/metric_interpretation/main/code/custom_functions_final.R')

```


## Sidebar {.sidebar}

<h5><b>Required Inputs</b></h5>


```{r shiny_inputs1}
numericInput(
  'sim_num',
  label = 'Simulation Replications:',
  value = 10^3,
  min = 10^3,
  max = 10^5,
  step = 1000,
  width = NULL
)

numericInput(
  'num_classes',
  label = 'Number of Classes:',
  value = 2,
  min = 2,
  max = 20,
  step = 1,
  width = NULL
)


numericInput(
  'num_observations',
  label = 'Number of Observations:',
  value = 100,
  min = 100,
  max = 10^5,
  step = 1,
  width = NULL
)


numericInput(
  'first_class_perc',
  label = 'Proportion of Reference Class:',
  value = 0.2,
  min = 0.001,
  max = 0.999,
  step = NA,
  width = NULL
)


selectInput(inputId = 'pred_approach', 
            label = 'Prediction Approach:',
            choices = c('random', 'proportional'),
            selected = 'proportional')


numericInput(
  'quantile',
  label = 'Quantile Computed for Classification Metrics:',
  value = 0.99,
  min = 0.005,
  max = 0.995,
  step = 0.005,
  width = NULL
)

```



---

<h5><b>Development Team</b></h5>  
[Fadel M. Megahed](https://miamioh.edu/fsb/directory/?up=/directory/megahefm)   
[Allison Jones-Farmer](https://miamioh.edu/fsb/directory/?up=/directory/farmerl2)     
[Steven E. Rigdon](https://www.slu.edu/public-health-social-justice/faculty/rigdon-steven.php)

**Version:** 0.1.0


---

<h5><b>For Correspondence</b></h5>  
`r fa("inbox", fill = "steelblue")` $~$ [fmegahed\@miamioh.edu](mailto:fmegahed@miamioh.edu)  
`r fa("twitter", fill = "steelblue")` $~$ [\@FadelMegahed](https://twitter.com/fadelmegahed)  
`r fa("github", fill = "steelblue")` $~$ [\@fmegahed](https://github.com/fmegahed)  



## Summary Column
  
### Summary of Simulation Results
  
```{r simulation}

# The five experimental conditions:
# _________________________________

input_sim_num = reactive( x = 1:as.numeric(input$sim_num) )
input_num_classes = reactive( x = as.numeric(input$num_classes) )
input_num_observations = reactive( x = as.numeric(input$num_observations) )
input_first_class_perc = reactive( x = as.numeric(input$first_class_perc) )
input_pred_approach = reactive(x = as.character(input$pred_approach))


sim_setup = reactive({
  sim_grid = expand_grid(
    input_sim_num(), 
    input_num_classes(), 
    input_num_observations(), 
    input_first_class_perc(),
    input_pred_approach()
  ) 
  
  colnames(sim_grid) = c('sim_num', 'num_classes', 'num_observations',
                         'first_class_perc', 'pred_approach')
  
  sim_grid %>%
    group_by_at( vars(-sim_num ) ) %>% 
    # using the dplyr::cur_group_id() to create the IDs
    mutate( exp_num = cur_group_id() ) %>% 
    # ungrouping since it was only performed to create unique ids
    ungroup() %>% 
    # moving the exp_num to the beginning
    relocate( exp_num )  %>% 
    # creating list columns for all the columns with except of ID-type columns
    nest(data = any_of( c('num_classes', 'num_observations',
                         'first_class_perc', 'pred_approach') ) ) %>% 
    arrange( exp_num ) -> sim_setup
  
  return(sim_setup)
})

# reactive(write_rds( sim_setup(), 'sim_setup.rds'  ))


experimental_results = reactive({
  results = sim_setup() %>% 
  mutate(
    acc_metrics = map(.x = data, .f = sim_function, base_class_name = 'class_')
  )
  
  return(results)
})

# reactive(write_rds( experimental_results(), 'experimental_results.rds'  ))


summary_results = reactive({
   experimental_results() %>% 
  # unnesting the acc_metrics (i.e., generating columns for each of the metrics)
  unnest( acc_metrics ) %>% 
  # grouping by exp_num 
  # (and data which is redundant but to keep it in the summary table)
  group_by(exp_num, data) %>% 
  # creating summaries of means and sds for each of the metrics by exp_num
  summarise(
    across(accuracy:g_mean,  
           list(mean = ~ mean(.x, na.rm = TRUE),
                sd = ~ sd(.x, na.rm = TRUE), 
           quantile = ~ quantile(.x, probs = as.numeric(input$quantile), na.rm = TRUE) ),
           .names = "{.fn}_{.col}" )
  ) %>% 
  # expanding the data into the different experimental conditions for interpretation
  unnest(data)
})

# reactive(write_rds( summary_results(), 'summary_results.rds'  ))
 


renderUI({
  
  results = summary_results()
  
  line1 = '<h4><b>Your selected inputs:</b></h4>'
  line2 = paste0('<b>Number of simulations:</b>  ', scales::comma( max( input_sim_num() ) ) )
  line3 = paste0('<b>Number of classes:</b>  ', input_num_classes())
  line4 = paste0('<b>Number of observations:</b>  ', scales::comma(input_num_observations()) )
  line5 = paste0('<b>Reference class proportion:</b>  ', input_first_class_perc())
  line6 = paste0('<b>Prediction approach:</b>  ', input_pred_approach())
  line7 = ' <br> '
  line8 = '<h4><b>A summary of the simulation results:</b></h4>'
  
  line9 = paste0(
    '<b>Accuracy:</b> ',
    round(results$mean_accuracy, digits = 3) %>% format(nsmall = 3),
    ' +/- ',
    round(results$sd_accuracy, digits = 3) %>% format(nsmall = 3),
    ' with ',
    as.numeric(input$quantile) * 100,
    '% of the accuracy metric values <= ',
    round(results$quantile_accuracy, digits = 3) %>% format(nsmall = 3)
  )
  
  line10 = paste0(
    '<b>Sensitivity:</b>  ',
    round(results$mean_sensitivity, digits = 3) %>% format(nsmall = 3),
    ' +/- ',
    round(results$sd_sensitivity, digits = 3),
    ' with ',
    as.numeric(input$quantile) * 100,
    '% of the accuracy metric values <= ',
    round(results$quantile_sensitivity, digits = 3) %>% format(nsmall = 3)
  )
  
  
  line11 = paste0(
    '<b>Specificity:</b>  ',
    round(results$mean_specificity, digits = 3) %>% format(nsmall = 3),
    ' +/- ',
    round(results$sd_specificity, digits = 3) %>% format(nsmall = 3),
    ' with ',
    as.numeric(input$quantile) * 100,
    '% of the accuracy metric values <= ',
    round(results$quantile_specificity, digits = 3) %>% format(nsmall = 3)
  )
  
  line12 = paste0(
    '<b>f_measure:</b>  ',
    round(results$mean_f_measure, digits = 3) %>% format(nsmall = 3),
    ' +/- ',
    round(results$sd_f_measure, digits = 3) %>% format(nsmall = 3),
    ' with ',
    as.numeric(input$quantile) * 100,
    '% of the accuracy metric values <= ',
    round(results$quantile_f_measure, digits = 3) %>% format(nsmall = 3)
  )
  
  line13 = paste0(
    '<b>G-mean:</b>  ',
    round(results$mean_g_mean, digits = 3) %>% format(nsmall = 3),
    ' +/- ',
    round(results$sd_g_mean, digits = 3) %>% format(nsmall = 3),
    ' with ',
    as.numeric(input$quantile) * 100,
    '% of the accuracy metric values <= ',
    round(results$quantile_g_mean, digits = 3) %>% format(nsmall = 3)
  )
  
  
  HTML(paste(line1, line2, line3, line4, line5, line6, line7,
             line8, line9, line10, line11, line12, line13, 
             sep = '<br/>'))
}
)



```


## Plots {.tabset .tabset-fade .tabset-pills}

### Overall Accuracy

```{r acc_plot}
mean_acc = reactive({
    experimental_results() %>% 
    # unnesting the acc_metrics (i.e., generating columns for each of the metrics)
  unnest( acc_metrics ) %>% 
    select(-data) %>% 
    pivot_longer(cols = accuracy:g_mean, names_to = 'metric') %>% 
    ungroup() %>% 
    group_by(metric) %>% 
    summarise(mean = mean(value, na.rm = T) %>% round(digits = 3)) %>% 
    filter(metric == 'accuracy')
})

plot_data = reactive({
  experimental_results() %>% 
    # unnesting the acc_metrics (i.e., generating columns for each of the metrics)
  unnest( acc_metrics ) %>% 
    select(-data) %>% 
    pivot_longer(cols = accuracy:g_mean, names_to = 'metric')
})


renderPlotly({
  plot_data() %>% 
    filter(metric == 'accuracy') %>% 
    ggplot(aes(x = value, group = metric)) +
  geom_histogram(aes(fill = as.factor(metric)), binwidth = 0.01, color = 'black') +
  geom_vline(data = mean_acc(), 
             aes(xintercept = mean), col='black', size = 1.75, linetype = 'dotdash') + 
  scale_x_continuous(breaks = scales::pretty_breaks(10), limits = c(0,1)) +
  scale_y_continuous(breaks = scales::pretty_breaks(5), labels = scales::comma) +
  scale_fill_manual(values = '#E31A1C') +
  scale_color_manual(values = '#FB9A99') +
  theme_bw() +
  theme(legend.position = 'none',
        plot.title = element_text(size=11)) + 
  labs(x = 'Accuracy Values', y = 'Count', 
       title = paste('Values based on',
                     scales::comma(input_sim_num() %>% max()), 'simulations,',
                     'with a reference class proportion of', 
                     input_first_class_perc(), 'and ',
                     paste0('"', input_pred_approach(), '"'), 'predictions')) -> p
  
  plotly::ggplotly()
})


```



### Histogram of ByClass Metrics

```{r histogram_plot}
df_mean = reactive({
  experimental_results() %>% 
    # unnesting the acc_metrics (i.e., generating columns for each of the metrics)
  unnest( acc_metrics ) %>% 
    select(-data) %>% 
    pivot_longer(cols = accuracy:g_mean, names_to = 'metric') %>% 
    ungroup() %>% 
    group_by(metric) %>% 
    summarise(mean = mean(value, na.rm = T) %>% round(digits = 3)) %>% 
    filter(metric != 'accuracy')
})


renderPlotly({
  plot_data() %>% 
    filter(metric != 'accuracy') %>% 
    ggplot(aes(x = value, group = metric)) +
    facet_wrap(~ metric, scales = 'fixed') +
  geom_histogram(aes(fill = as.factor(metric)), binwidth = 0.01, color = 'black') +
  geom_vline(data = df_mean(), 
             aes(xintercept = mean), col='black', size = 1.75, linetype = 'dotdash') + 
  scale_x_continuous(breaks = scales::pretty_breaks(10), limits = c(0,1)) +
  scale_y_continuous(breaks = scales::pretty_breaks(5), labels = scales::comma) +
  scale_fill_brewer(palette = "Paired") +
  scale_color_brewer(palette = "Paired") +
  theme_bw() +
  theme(legend.position = 'none',
        plot.title = element_text(size=11)) + 
  labs(x = 'Metric Values', y = 'Count', 
       title = paste('Values based on',
                     scales::comma(input_sim_num() %>% max()), 'simulations,',
                     'with a reference class proportion of', 
                     input_first_class_perc(), ' & ',
                     paste0('"', input_pred_approach(), '"'), ' predictions')) -> p
  
  plotly::ggplotly()
})
```

